Stack Safety as a Security Property

### 2020-07-09 main.tex
Polishing

The call stack is a perennial target for attacks
based on memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \apt{Mostly from
  nick;there could be more}\bcp{How about static techniques??  We decided
  these were also in-scope.}\bcp{... or are they?  I guess in any case we should
list any that we know about.}

But what {\em is} this ``stack safety'' of which we speak? Precise formal
definitions are lacking in the literature; instead, the notion
is usually defined negatively, by enumerating the bad things that a given
enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one---quite recent---attempt to frame a
positive definition:
\citet{Skorstengaard+19} define stack safety as
the conjunction of two properties, {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.''
%
Although this represents an important step in the right direction, we believe
there is still room to craft a more abstract, more generic characterization
of stack safety that is independent of capabilities or any other specific
enforcement
technique.  In other words, we seek formal criteria that can be applied
against \emph{any} stack-based system + enforcement mechanism to judge
whether or not the combination is actually stack-safe.

\iflater
\leo{Should this be its own paragraph?}
\fi
The key contribution of this paper is a novel, generic presentation of
local state encapsulation and well-bracketed control flow using the
technical framework of language-based security.

The intuition behind \citeauthor{Skorstengaard+19}'s use of the term {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should not be able to access or overwrite private data
in the caller's stack frame. We formalize this intuition as a security
property with both confidentiality and integrity components.
%
We reflect this intuition in a formal
{\em stack confidentiality} property, which promises that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g., a caller's private data or uninitialized memory).  Conversely, {\em stack
  integrity} promises that, if a function call ever returns, the caller's
stack data at the point of return will be identical to what it was before
the call.

@@

The intuition behind \citeauthor{Skorstengaard+19}'s {\em well-bracketed control
  flow} is
the desire to protect the call graph itself: a callee should pass control
back to the caller only by returning in the standard way (transferring
control to the instruction following the call); it should
not be
allowed to jump back into the caller's code at an arbitrary point, either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety
between functions: each instruction is uniquely owned by some
function, and all jumps between functions are either calls that enter
at a marked entry point or returns that re-enter just after their
associated call in the standard nested call structure.  \apt{clarify
  what we are contributing here.}

@@

\leo{We've said this literally in the previous sentence. What should this intro
  be in the new reorg?}
%We begin with a running example that will serve both to demonstrate several
%``stack-unsafe'' behaviors that our stack safety property should prohibit
%and to ground the technical presentation in later sections.

The example program appears in \Cref{fig:running-program}. It consists
of a {\tt main} function, written in some C-like high-level language,
that performs a call to some function {\tt f}, which in turns makes a
nested call to some function {\tt g} before returning, while
performing some simple arithmetic operations in the process.

@@

To show a trace of the execution of this program, we need a low-level
view of both the code and the machine it runs on. While the concrete
implementation of our enforcement mechanism (\Cref{sec:enforcement})\apt{Is it ``ours'' in this context?}
operates on a full conventional RISC-V architecture (using the FORVIS
simulator\leo{cite}\apt{why mention this here?}), we make a few simplifying assumptions here to
ease the exposition.

First of all, we ignore the heap. We are mainly concerned with stack
interactions so that will be the main focus; integrating a safety
propety for the heap in the
style of \citet{AzevedoDeAmorim17} is a natural extension we discuss
in~\Cref{sec:future}.
%
We also abstract away low-level details of memory layout and present a
machine that is word-addressed, with instructions starting from
location 0, a dedicated memory-mapped output port at location 99, and
the stack beginning at location 100 and growing upwards with increasing
addresses.  \leo{Explain more? How to connect to realistic
  one? Observations? DMA? Interrupt handlers?}
%
A sketch of the initial memory layout can be found at the top of
\cref{fig:running-trace-b}: instructions for {\tt main} begin at 0,
the dedicated output is uninitialized, while the (also uninitialized)
stack begins at 100. Directly below the layout, we summarize the parts
of the register file that are relevant to the example.
%
The rest of \cref{fig:running-trace-b} traces the execution of the
program, showing the machine state at call and return points.
Instructions are intended to
be similar to RISC-V, but are described in an ad-hoc notation
rather than using assembler syntax. The $\jal$ (jump-and-link)
instruction used for calls atomically saves the return address in a register
and jumps to a specified new $\PCname$; the $\jalr$ (jump-and-link-via-register)
instruction used for returns jumps to the address stored in a register.
Register $r_0$ is hardwired to zero. \leo{Expand upon this?}

>>>>

\bcp{We don't actually say very clearly anywhere that we are interested in
  low-level enforcement mechanisms for machine-code programs (because this
  is where stack attacks occur, and because at higher levels of abstraction
  the stack is a built-in concept, not something that can be attacked even
  in principle).  This could cause confusion.}

The call stack is a perennial target for attacks
based on memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \iflater\apt{Mostly from
  nick;there could be more}\bcp{How about static techniques??  We decided
  these were also in-scope.}\bcp{... or are they?  I guess in any case we should
list any that we know about.}\fi

But what {\em is} this stack safety of which people speak? Precise formal
definitions are mostly lacking: rather, the notion
is commonly defined in a negative way, by enumerating the bad things that a
given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables\bcp{Does use of unitialized variables lead to
stack safety attacks?}.
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a
positive definition, by
\citet{Skorstengaard+19}, who define stack safety as
the conjunction of two properties: {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this represents an important step in the right direction, we believe
there is still room to craft a more abstract, more generic characterization
of stack safety that is independent of capabilities or any other specific
enforcement
technique.  In other words, we seek formal criteria that can be applied
to \emph{any} mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.''
%
Our key contribution is a novel, generic presentation of
local state encapsulation and well-bracketed control flow using the
technical framework of language-based security.

The intuition behind \citeauthor{Skorstengaard+19}'s use of the term {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should not be able to access or overwrite private data
in the caller's stack frame.
%
We formalize this intuition in (1) a {\em stack confidentiality}
property, which promises that a callee's observable behavior is not affected
by memory outside its stack frame (e.g., a caller's private data or
uninitialized memory), plus (2) a {\em stack integrity} property,
which promises that, if a function call ever returns, the caller's stack
data at the point of return will be identical to what it was before the
call.

@@

The intuition behind \citeauthor{Skorstengaard+19}'s {\em well-bracketed control
  flow} is
the need to protect the call graph itself: a callee should pass control
back to the caller only by returning in the standard way (transferring
control to the instruction following the call); it should
not be
allowed to jump back into the caller's code at an arbitrary point, either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety
between functions: each instruction is uniquely owned by some
function, and all jumps between functions are either calls that enter
at a marked entry point or returns that re-enter just after their
associated call in the standard nested call structure.  \apt{clarify
  what we are contributing here.}

@@

@@

\Cref{fig:running-program} shows a high-level view of our running example,
written in C-like pseudo-code.  It consists of a {\tt main} function that
performs a call to some function {\tt f}, which in turns makes a nested call
to some function {\tt g} before returning, while performing some simple
arithmetic operations in the process. \cref{fig:running-trace-b} shows an
execution trace of corresponding machine instructions. (These might have
been generated by compiling the high-level code, but it is only the behavior
of the machine instructions that we care about
here---\Cref{fig:running-program} is our actual running example.)

To streamline the exposition of the example, we use a small instruction set
reminiscent of RISC-V and make few simplifying assumptions.
%
First, since our goal is to formalize stack safety, we ignore the heap and
assume that the memory is divided between the instructions and the stack.
(Integrating a memory safety property for the heap in the style of
\citet{AzevedoDeAmorim17} is a natural extension that we discuss
in~\Cref{sec:future}.)
%
We also abstract away low-level details of memory layout and assume a
machine that is word- (not byte-) addressed, with instructions starting from
location 0, a dedicated memory-mapped output port at location 99, and
the stack beginning at location 100 and growing upwards with increasing
addresses.  \leo{Explain more? How to connect to realistic
  one? Observations? DMA? Interrupt handlers?}\bcp{Yes.}

A sketch of the initial memory layout appears at the top of
\cref{fig:running-trace-b}: the instructions for {\tt main} begin at 0,
the dedicated output is uninitialized, while the (also uninitialized)
stack begins at 100. Directly below the layout, we summarize the parts
of the register file that are relevant to the example.
%
The rest of \cref{fig:running-trace-b} traces the execution of the
program, showing the machine state at call and return points.
Instructions are intended to
be similar to RISC-V, but are described in an ad-hoc notation
rather than using assembler syntax. The $\jal$ (jump-and-link)
instruction used for calls atomically saves the return address in a register
and jumps to a specified new $\PCname$; the $\jalr$ (jump-and-link-via-register)
instruction used for returns jumps to the address stored in a register.
Register $r_0$ is hardwired to zero. \leo{Expand upon this?}\bcp{Yes, if it
  matters to understanding the example---point to where.}
